---
title: "CS-201 First Look at R"
author: "Amber Camp"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

### Load Library

```{r}
library(ggplot2)


```

### Some Scripting

```{r}
# This is a comment
Quiz_1 <- 88.  # This is how you assign a value to a nmae
Quiz_2 <- 94
Quiz_3 <- 79


Quiz_1


# averaging

(Quiz_1 + Quiz_2 + Quiz_3) / 3

# mean function 

mean(Quiz_1 + Quiz_2 + Quiz_3) # Not correct, just adds everything up 
mean(Quiz_1, Quiz_2, Quiz_3) # Not correct, it just takes the mean of the first value 
mean(c(Quiz_1, Quiz_2, Quiz_3)) # Correct!

# List 

list <- c(Quiz_1, Quiz_2, Quiz_3)

mean(list)

average <- mean(list)

# Assigning New value
Quiz_3 <- 99
Quiz_3

average

list <- c(Quiz_1, Quiz_2, Quiz_3)
average <- mean(list)
average

# Round a number 
round(average, 2)
```

### More Scores

```{r}
scores <- c(88,94,79,91,85)
scores

mean(scores)
min(scores)
max(scores)
length(scores)

summary(scores)

students <- c("a", "b", "c", "d", "e")
students

#barplot
barplot(scores, name.arg = students)

```

## Introducing `ggplot2`

```{r}

```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}

```

What did you observe?

> Answer:

Why do you think this is happening? (You can look online for help)

> Answer:
